% Section: Algebra and Coalgebra: Two Faces of Structure

Within category theory, a fundamental duality emerged: \textbf{algebra} and \textbf{coalgebra}.

\subsection{Algebra: Building Up}

An \textbf{algebra} for a functor $F$ is a structure where you can ``fold'' or ``evaluate'' $F$-structured data.

\begin{example}[Lists as an algebra]
The type of lists of natural numbers satisfies:
\[
\mathsf{List}(\mathbb{N}) \cong 1 + \mathbb{N} \times \mathsf{List}(\mathbb{N})
\]
A list is either empty ($1$) or a head ($\mathbb{N}$) followed by a tail (another list).

This is an \textbf{algebra} for the functor $F(X) = 1 + \mathbb{N} \times X$.

Functions on lists (like $\mathsf{sum}$) are defined by \textbf{folding}: specify what to do with empty and what to do with cons.
\end{example}

Algebras are about \textbf{construction} and \textbf{induction}: building finite, well-founded structures.

The syntax of a logic---formulas, proofs---forms an algebra: the \textbf{initial algebra} of the grammar.

\subsection{Coalgebra: Observing Behavior}

A \textbf{coalgebra} for a functor $F$ is a structure where you can ``unfold'' or ``observe'' behavior.

\begin{example}[Streams as a coalgebra]
An infinite stream of natural numbers has type:
\[
\mathsf{Stream}(\mathbb{N}) \to \mathbb{N} \times \mathsf{Stream}(\mathbb{N})
\]
From a stream, you can observe: the head (a number) and the tail (another stream).

This is a \textbf{coalgebra} for the functor $F(X) = \mathbb{N} \times X$.

Functions producing streams are defined by \textbf{unfolding}: specify the head and how to continue.
\end{example}

Coalgebras are about \textbf{behavior} and \textbf{coinduction}: potentially infinite, observable processes.

State machines, automata, transition systems---these are coalgebras.

\subsection{The Duality}

\begin{center}
\begin{tabular}{lll}
& \textbf{Algebra} & \textbf{Coalgebra} \\
\hline
Perspective & Construction & Observation \\
Principle & Induction & Coinduction \\
Structure & Finite, well-founded & Potentially infinite \\
Examples & Syntax, data types & Behavior, state machines \\
Canonical object & Initial algebra & Final coalgebra \\
\end{tabular}
\end{center}

\begin{keyinsight}
Algebra is about \textbf{syntax}: how things are built.\\
Coalgebra is about \textbf{semantics}: how things behave.

Logic lives in both: the formulas are algebraic (syntax), but their meaning involves coalgebraic structures (models, possible worlds, state transitions).
\end{keyinsight}
